Reviving the Lost Art of Combinator Graph Reduction

### 2023-01-24 experiments/superg/doc/paper/paper.tex
Write conclusion.

@@

\section{Conclusion}
% Repeat the main question / objective of your research

% Give the answer to the question
Source code for our Superg evaluator is available at \url{https://github.com/wildarch/mono/tree/main/experiments/superg}.

% Why is this the answer? Present the main arguments

% What are the implications of this answer?

% Suggestions for future research

>>>>

While the $\linear$ strategy produces slightly more compact code than $\strict$ and $\lazy$, there is no clear performance improvement.
Our small benchmark programs only require bulk combinators for up to $2$ arguments, so the $\linear$ strategy cannot win much in terms of code size.
We expect that for certain worst-case inputs, as given in~\cite{kiselyov_lambda_2018}, $\linear$ may still prove useful.

@@

\section{Conclusion}
% Repeat the main question / objective of your research
Our objective with this work was to revive research around Combinator Graph Reduction by improving the performance.
We have attempted this by using a novel semantic translation algorithm to compile lambda calculus programs into combinator expressions~\cite{kiselyov_lambda_2018}, and by implementing combinator graph reduction engines based on the designs of Miranda~\cite{turner_new_1979} and TIGRE~\cite{koopman_fresh_1989}.

% Give the answer to the question
We have developed a reference implementation of a combinator graph reduction engine based on the now open source Miranda~\cite{turner_new_1979}.
% Why is this the answer? Present the main arguments
Our implementation comprises under 1000 lines of Rust code and outperforms the original Miranda engine on our benchmarks.

% Give the answer to the question
Our port of the TIGRE~\cite{koopman_fresh_1989} engine runs natively on stock x86-64 hardware and executes our benchmark program with fewer instructions and data accesses than our Miranda-style engine.
% Why is this the answer? Present the main arguments
Unfortunately, its performance still lags behind the reference implementation because its inherent frequent writes to executable code sections trigger expensive invalidations of the instruction cache.
We conclude that the TIGRE design is a poor fit for architectures with split instruction and data caches, though some concepts like \texttt{LIT} combinators and tag elimination may still prove effective.

% Give the answer to the question
Applying Kiselyov's translation algorithms~\cite{kiselyov_lambda_2018} to our benchmarks was a resounding success.
% Why is this the answer? Present the main arguments
All proposed variations of the semantic translation produced a combinator expression that was smaller than the one generated by the reference bracket abstraction algorithm.
Moreover, the effect of the smaller code size was amplified at execution time, with the smallest program running nearly twice as fast as the reference program.

% Suggestions for future research
There is ample room for improvement both on the compilation and the runtime side.
Our $\linear$ translation, like Kiselyov's, is based on the $\strict$ translation rather than the $\lazy$ or $\lazyeta$ translation schemes.
Combining bulk combinators with lazy weakening could further reduce the size of compiled programs, and therefore also improve performance.

On the runtime side, ideas from TIGRE could be selectively integrated into the Miranda-style engine to improve its performance, such as \texttt{LIT} combinators or eliminating tags.

Our current runtime is single-threaded: it always reduces one combinator at a time.
However, reducing multiple combinators in parallel still leads to a correct execution, so it may be possible to run programs faster by evaluating separate parts of the graph in parallel on different cores.
As suggested by Kmett~\cite{kmett_combinators_2018}, it may even be possible to use SIMD instructions to reduce multiple combinators in parallel on a \emph{single} core.

Our garbage collector and the one used by Miranda is a simple mark-and-sweep collector.
While our benchmark programs do not run long enough to exhaust the available memory, we expect that for real programs a faster garbage collector could improve performance significantly.
In particular, a concurrent garbage collector~\cite{jones_garbage_2011} could prevent execution stalls during collection time.
We hypothesize that a limited form of reference counting would also be beneficial: the S combinator (B and C are very similar) currently requires the allocation of two new nodes when it executes, and it also drops a reference to two other nodes.
With reference counting we could determine if those two nodes are not needed elsewhere in the graph, and reuse them, eliminating the creation of garbage.

Source code for our translation algorithms and runtimes, as well as associated tests and benchmarks, are available at \url{https://github.com/wildarch/mono/tree/main/experiments/superg}.
